A general Nullstellensatz for generalized spaces

### 2019-03-16 paper-qcoh.tex
typo

\begin{abstract}
  We give a general Nullstellensatz for the generic model of a
  geometric theory, useful as a source of nongeometric sequents validated by
  the generic model, and characterize the first-order formulas validated by the
  genetric model. XXX
\end{abstract}

@@

\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
of rings, local rings or intervals. We follow Caramello's
terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
given by a set of sorts, a set of finitary function symbols, a set of finitary
relation symbols and a set of axioms, consisting of geometric sequents
(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
are geometric formulas, that is formulas built from equality and the relation
symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$). By \emph{first-order
formula} we will mean a formula which may contain, additionally to the
connectives allowed for geometric formulas, the connectives~${\Rightarrow}$
and~${\forall}$.

A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
a model such that for any geometric sequent~$\sigma$, the following notions
coincide:
\begin{enumerate}
\item The sequent~$\sigma$ is provable modulo~$\TT$.
\item The sequent~$\sigma$ holds for any~$\TT$-model in any Grothendieck topos.
\item The sequent~$\sigma$ holds for~$U_\TT$.
\end{enumerate}
One could argue that it is this model which mathematicians
implicitly refer to when they utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
typically not be realized as a set-theoretic model, consisting of a set for
each sort, a function for each function symbol and so on; instead it is a model
in a custom-tailored syntactically constructed Grothendieck topos, the
\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.

To state what it means for a~$\TT$-structure in a topos~$\E$ to verify the
axioms of~$\TT$, rendering it a model, the \emph{internal language} of~$\E$ is
used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
\models \alpha$'' to mean that a formula~$\alpha$ holds from the internal
point of view of~$\E$. Since this language is a form of a higher-order
intuitionistic extensional dependent type theory, the classifying topos~$\Set[\TT]$ can
be regarded as a higher-order completion of the geometric theory~$\TT$. The
generic model enjoys the universal property that any~$\TT$-model in any
(Grothendieck) topos~$\E$ is the pullback of~$U_\TT$ along an essentially
unique geometric morphism~$\E \to \Set[\TT]$.
\medskip

@@

Let~$A$ be a ring. A simple version of the classical Nullstellensatz states:
\emph{For any polynomials~$f$ and~$g$ over~$A$, if any zero of~$f$ is also a
zero of~$g$, then there is a polynomial~$h$ such that~$g = hf$.} The
polynomial~$h$ can be regarded as an ``algebraic certificate'' of the
hypothesis. This principle holds for instance in the case that~$A$ is an
algebraically closed field and that~$g$ is the unit polynomial. We will see
below that it is also
true, without any restriction on~$g$, for the generic ring.

@@

To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
\[ \Set[\TT] \models
  \forall f,g \? U_\TT[X]\_ \bigl(
    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
      \exists h \in U_\TT[X]\_ g = hf\bigr). \]

The statement~\eqref{eq:nullstellensatz} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. However, there is still a useful substitute,
which formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
substantially broadens the scope of the Nullstellensatz.

@@

\begin{thm}\label{thm:characterization0}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom
scheme~\eqref{eq:nullstellensatz}.
% XXX be more precise about "scheme"
\end{enumerate}
\end{thm}

@@

@@

\subsection{Background on the internal language of Grothendieck toposes}

@@

\begin{proof}The direction~$(1) \Rightarrow (2)$ is immediate, as the internal
language of~$\Set[\TT]$ validates first-order intuitionistic logic and as it
validates the
axiom scheme~\eqref{eq:nullstellensatz} by Theorem~\ref{thm:nullstellensatz}.

@@

\begin{lemma}The functor~$(\cdot)^\sim$ from~$\TT$-models to~$\ul{\TT}/U_\TT$-models
in~$\Set[\TT]$ is left adjoint to functor~$\Gamma = \Hom(1, \cdot)$ computing
global elements.
\end{lemma}

@@

% XXX fill in proof

@@

* Cite https://docplayer.net/55796150-Intuitionistic-algebra-some-recent-developments-in-topos-theory.html
  (in general and for internal theories)

* Discuss related work, especially: Olivia 6.1.3, ...

% We would not expect the statement
% \[ \text{``for any geometric sequent~$\sigma$, if~$\sigma$ holds for~$A$
% then the theory of rings proves~$\sigma$''} \]
% to hold for an ordinary ring~$A$ in the topos of sets.
%
% Let~$M$ be a~$\TT$-model. We then trivially have
% signature of~$\TT$. While we always have the implication
% \[ \text{$\TT$ proves~$\sigma$} \quad\Longrightarrow\quad
%   \text{$\sigma$ holds for~$M$}, \]
% the converse is of course typically false,
%
% Furthermore, in a first introduction, it is useful to
% restrict to an instructive special case, such as the theory of (commutative) rings.
%
% A simple version of the classical Nullstellensatz, valid for algebraically
% closed fields, states: Let~$f \in A[X]$ be a polynomial. If
% \[ \forall x \in A\_ (f(x) = 0 \in A \Rightarrow 1 = 0 \in A), \]
% then there is a polynomial~$g \in A[X]$ such that~$gf = 1$. This polynomial~$g$
% can be regarded as an ``algebraic certificate'' that the equation~$f(x) = 0$
% only has a solution in the trivial case~$1 = 0$. The conclusion can also be phrased as
% follows: In the free~$A$-algebra on one generator modulo the relation~$f(X) =
% 0$, the relation~$1 = 0$ holds.
%
% Let~$A$ be a specific ring. Let~$f,g \in \ZZ[X]$ be
% polynomials. While we always have the implication
% \[ g([X]) = 0 \in \ZZ[X]/(f) \quad\Longrightarrow\quad
%   \forall x \in A\_ (f(x) = 0 \Rightarrow g(x) = 0), \]
% the converse is typically false: Just that~$g(x) = 0$ holds for every zero
% of~$f$ in~$A$ does not imply that~$g([X]) = 0$ holds for the generic zero
% of~$f$ in the free ring on one generator~$X$ modulo the relation~$f(X) = 0$.
%
%
% the case that~$\TT$ is a Horn theory (where the axioms are of the
% form~$(\phi \seq{\vec x} \psi)$ with the formulas~$\phi$ and~$\psi$ being
% finite conjunctions of atomic propositions (equality and the given relations)),
% such that we can switch from a logical terminology to a more algebraic one.
% In the Horn case, for any set~$X$ and any set~$R$ of atomic propositions the
% term algebra construction yields~$\TT\langle X | R\rangle$, the
% free~$\TT$-model generated by~$X$ modulo the relations~$R$. We will also abuse
% notation to pretend that~$\TT$ is single-sorted.
%
% Let~$M$ be a specific~$\TT$-model and assume that for
% all~$n$-tuples~$x = (x_1,\ldots,x_n)$ of elements of~$M$, if some set~$R_1$ of
% atomic propositions is satisfied for~$x$ then so is some set~$R_2$. In this
% case it is unreasonable to expect that~$R_2$ holds for the~$n$-tuple of
% generators of~$\TT\langle e_1,\ldots,e_n | R_1\rangle$.
%
% let~$R_1$ and~$R_2$ be sets of atomic
% propositions with~$n$ free variables. Then
% and assume, for some sets of relations~$R_1$
% and~$R_2$
%
% The term algebra of a Horn theory~$\TT$ is always a model of~$\TT$, such that

* Sei T' subkanonisch für T.

* Sei φ eine Hornformel. Gelte ∇φ. Folgt dann φ?

* Für φ = (x = y) stimmt das, da U separiert ist.

* Für φ = (R(x,y)) stimmt, das ...?

  Was heißt das denn? Habe R \--> U x U. Gehe davon aus, dass ∇φ bedeutet, dass
  ∇((x,y) ∈ R). Daraus folgt aber (x,y) ∈ R:

  K = { (x,y) | (x,y) ∈ R } ist Subsingleton von U x U und ∇-bewohnt.
  Also existiert (a,b) ∈ U x U mit ∇... Hilft nicht.

  Aber: K = { (a,b) ∈ R | a = x & b = y } ist Subsingleton von R.
  Und ∇-bewohnt. Also existiert (a,b) ∈ R mit ∇(a=x & b = y).
  Wegen Separiertheit folgt a = x & b = y. Fertig. :-)

@@

* Nichttrivialität des Nullstellensatzes diskutieren (Spezialfall top |- φ
  betrachten, dieser ist nämlich schon mehr oder weniger trivial)

* Anmerken, dass der Nullstellensatz für first-order sequents NICHT stimmt
  (Beispiel: ¬(x=0) ⇒ (x inv.))

* Anmerken, dass wir in konstruktiver Metalogik arbeiten.

>>>>

\begin{abstract}
  We give a general Nullstellensatz for the generic model of a
  geometric theory, useful as a source of nongeometric sequents validated by
  the generic model, and characterize the first-order and higher-order formulas validated by the
  genetric model.
\end{abstract}

@@

\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
of rings, of local rings or of intervals. We follow Caramello's
terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
given by a set of sorts, a set of finitary function symbols, a set of finitary
relation symbols and a set of axioms, consisting of geometric sequents
(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
are geometric formulas, that is formulas built from equality and the relation
symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$). By (infinitary) \emph{first-order
formula} we will mean a formula which may contain, additionally to the
connectives allowed for geometric formulas, the connectives~${\Rightarrow}$
and~${\forall}$.

A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
a model such that for any geometric sequent~$\sigma$, the following notions
coincide:
\begin{enumerate}
\item The sequent~$\sigma$ is provable modulo~$\TT$.
\item The sequent~$\sigma$ holds for any~$\TT$-model in any Grothendieck topos.
\item The sequent~$\sigma$ holds for~$U_\TT$.
\end{enumerate}
One could argue that it is this model which we
implicitly refer to when we utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
typically not be realized as a set-theoretic model, consisting of a set for
each sort, a function for each function symbol and so on; instead it is a model
in a custom-tailored syntactically constructed Grothendieck topos, the
\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.

To state what it means for a~$\TT$-structure in a topos~$\E$ to verify the
axioms of~$\TT$, rendering it a model, the \emph{internal language} of~$\E$ is
used, roughly reviewed in Section~\ref{sect:review-language} below. We write~``$\E
\models \alpha$'' to mean that a formula~$\alpha$ holds from the internal
point of view of~$\E$. Since this language is a form of a higher-order
intuitionistic extensional dependent type theory, the classifying topos~$\Set[\TT]$ can
be regarded as a higher-order completion of the geometric theory~$\TT$. The
generic model enjoys the universal property that any~$\TT$-model in any
(Grothendieck) topos~$\E$ is the pullback of~$U_\TT$ along an essentially
unique geometric morphism~$\E \to \Set[\TT]$.
\medskip

@@

Let~$A$ be a ring. A simple version of the classical Nullstellensatz states:
\emph{For any polynomials~$f$ and~$g$ over~$A$, if any zero of~$f$ is also a
zero of~$g$, then there is a polynomial~$h$ such that~$g = hf$.} The
polynomial~$h$ can be regarded as an ``algebraic certificate'' of the
hypothesis. This principle holds for instance in the case that~$A$ is an
algebraically closed field and~$g$ is the unit polynomial. We will see
below that it is also
true, without any restriction on~$g$, for the generic ring.

@@

To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
\[ \Set[\TT] \models
  \forall f,g \? U_\TT[X]\_ \bigl(
    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
      \exists h \? U_\TT[X]\_ g = hf\bigr). \]

The statement~\eqref{eq:nullstellensatz} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. However, there is still a useful substitute,
which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
substantially broadens the scope of the Nullstellensatz.

@@

\begin{thm}\label{thm:characterization0}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom~\eqref{eq:nullstellensatz}.
\end{enumerate}
\end{thm}

@@

% XXX outline and higher-order

@@

\subsection{Background on the internal language of Grothendieck toposes}
\label{sect:review-language}

@@

\begin{proof}The direction~$(2) \Rightarrow (1)$ is immediate, as the internal
language of~$\Set[\TT]$ validates first-order intuitionistic logic and as it
validates the
axiom scheme~\eqref{eq:nullstellensatz} by Theorem~\ref{thm:nullstellensatz}.

@@

\begin{lemma}The functor~$(\cdot)^\sim$ from~$\TT$-models to~$\ul{\TT}/U_\TT$-models
in~$\Set[\TT]$ is left adjoint to the functor~$\Gamma = \Hom(1, \cdot)$ computing
global elements.
\end{lemma}

@@

\begin{proof}Theorem~\ref{thm:definability} on the semantic side and the axiom
scheme on the syntactic side allow us to replace any mention of a
powersort~$P(X)$ in~$\alpha$
by~$\mathrm{Form}^\star_{x\!\?\!X}(\ul{\TT}/U_\TT)/({\dashv\vdash_x})$. Then we
can argue as in the proof of Theorem~\ref{thm:characterization}, noting that
the axiom scheme indeed entails the axiom scheme posited by
Theorem~\ref{thm:characterization}.\end{proof}

% XXX check

%In einem ersten Schritt schreiben wir alle Existenzquantifikationen und freie Variablen
%über Potenzsorten um. Aus "exists A : P(X)" wird "exists φ :
%Form*(T/U_T)_{x:X}/-||-". (Dabei kann X immer noch eine abgeleitete Sorte sein!) Aus
%der freien Variable "A : P(X)" wird analog eine freie Variable.
%
%Entsprechend verändern wir den Rumpf. Aus "x0 ∈ A" wird φ[x0/x].
%
%Dann sind wir fertig mit dem Nullstellensatz für extended geometric logic.
%
%Noch klären:
%* P(X) x Y
%* Form* Teil von extended geometric logic?
%* Beweisbarkeit modulo extended geometric logic ist geometrische Formel?

@@

* XXX Cite https://docplayer.net/55796150-Intuitionistic-algebra-some-recent-developments-in-topos-theory.html
  (in general and for internal theories)

* XXX Discuss related work, especially: Olivia 6.1.3, ...

@@

* XXX Nichttrivialität des Nullstellensatzes diskutieren (Spezialfall top |- φ
  betrachten, dieser ist nämlich schon mehr oder weniger trivial)

* XXX Anmerken, dass der Nullstellensatz für first-order sequents NICHT stimmt
  (Beispiel: ¬(x=0) ⇒ (x inv.))

* XXX Anmerken, dass wir in konstruktiver Metalogik arbeiten.
